---
title: "Quantifying Chaos"
author: "Murray Jones (520487532)"
date: "University of Sydney | PHYS1902 | Lab Class X"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    fig_caption: yes
    number_sections: yes
    self_contained: yes
    theme: flatly
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: hide
subtitle: Title
---

<br>

## Introduction:

### Aim:

The aim of this experiment is to determine a quantitative measure of the chaos in the motion of a triple pendulum, and use this to determine a relationship between the extent of chaos exhibited by the pendulum, and the the initial potential energy of the pendulum arm.

### Hypothesis:

The extent of chaos exhibited by the triple pendulum will be directly proportional to the initial energy of the pendulum.

### Measurable variables:

### Scientific basis:

Edward Lorenz defined chaos as:

> *When the present determines the future, but the approximate present does not approximately determine the future.*

Based on this definition, we can conclude that the extent of chaos expressed by a system over time can be measured as the difference in paths taken by a system over multiple trials whose initial conditions are approximately the same.

Double and triple pendula are classic examples of chaotic systems.

Initially, all the energy of the pendulum arm is in the form of gravitational potential energy. \

Refer to your measurable variables consistent with later analysis\
\~1-2 sentences of text

\
Scientific basis for study presented that reflects synthesis of research\
Including references to relevant journal papers and your own knowledge of\
physics

\
Include relevant equations and concepts\
Insightful context of the experiment and scientific basis\
What is the issue or problem your experiment addresses?

\
0.5 page

## Experimental Setup and Method:

Provide a list of equipment, and variables that were measured\
Describe your experimental method including calculating uncertainties\
Include diagrams where appropriate\
The reader should be able to reproduce your method\
Bullet points are OK, but use full sentences\
1-- 2 pages of text and diagram

## Data and Results:

Comment on the precision of your measurements. Indicate main sources of\
error, clearly explain how you estimated/calculated uncertainties.\
Present the final form of the analysed data (e.g. final graphs, tables, bar\
graphs, LINEST results). It should align with your aim.\
Include the method you used to analyse the data (e.g. I plotted these two\
variables against each other, I used Excel to fit this trend, and so on).

Don't forget to include units, uncertainties, and use correct number of\
significant figures.\
Describe your results\
Include tables of data and Excel plots where appropriate\
Make sure all data has units and appropriate uncertainties\
1-2 pages of text and diagrams

## Discussion and Interpretation:

Interpret the analysed data to answer the aim\
Comment on the validity, reliability and accuracy of the data\
Discuss your uncertainty analysis and sources of error.\
If there are discrepancies between predictions and measurements, discuss\
these and provide plausible explanations/justifications.\
Any limitations of your experimental approach?\
Connect your results to your experiment aims/hypotheses\
\~ 1 page of text

```{r, echo=FALSE}
data = read.csv("allData.csv")
```

```{r, echo=FALSE}
# Account for late starts:

isSameInitConds = function(is, as) {
    
    if (unique(is$Type)[1] == unique(as$Type)[1] &&
        unique(is$InitAngle)[1] == unique(as$InitAngle)[1] &&
        unique(is$AddedMass)[1] == unique(as$AddedMass)[1]) {
        
        return(1) # Same init conds.
    }
    
    return(0) # Not same init conds.
}

# Get an on time run with same initial conditions.
getOnTime = function(rec) {
    badRun = data[data$Recording == rec, ]
    
    for (rec in unique(data$Recording)) {
        
        testRun = data[data$Recording == rec, ]
        
        if(isSameInitConds(testRun, badRun)) {
        
            testLate = unique(testRun$LateStart)[1]
            
        
            if (testLate == 0) {
                return(testRun)
            }
        }
    }
}


shifts = c()

for (rec in unique(data$Recording)) {
    run = data[data$Recording == rec, ]
    
    late = unique(run$LateStart)[1]
    
    timeshift = 0
    
    if (late) {
        refRun = getOnTime(rec)
        
        target = run$GreenAngle[1]
    
        closePoints = refRun[abs(refRun$GreenAngle - target) < 0.4,]
        
        firstPoint = head(closePoints, 1)
    
        timeshift = firstPoint$Time

    }
    
    for (row in 1:nrow(run)) {
        shifts[length(shifts)+1] = timeshift
    }
}

data$Timeshifts = shifts

data$AdjTime = data$Time + data$Timeshifts

str(data)
```

```{r, echo=FALSE}
unique(data$InitAngle)
```

```{r, echo=FALSE}
data$GreenPos = data$GreenDist * data$GreenAngle

run = data[data$Recording == "268", ]
run2 = data[data$Recording == "266", ]
run3 = data[data$Recording == "267", ]


plot(run3$AdjTime, run3$GreenPos, cex=0,
     xlab="Time (s)", ylab="Angular displacement (m·°)")

lines(run$AdjTime, run$GreenPos, col="#114411")
lines(run2$AdjTime, run2$GreenPos, col="#11BB11")
lines(run3$AdjTime, run3$GreenPos, col="#110011")
```

```{r, echo=FALSE}
analysedRuns = c()


getSameCondRuns = function(source) {
    
    runs = c(source)
    
    #print("-------")
    
    for (rec in unique(data$Recording)) {
        
        if (rec == source) {
            next
        }
        
        if (isSameInitConds(data[data$Recording == rec, ],
                            data[data$Recording == source, ])) {
            
            
            runs[length(runs)+1] = rec
        }
    }
    
    return(runs)
}


difSet = c()
angSet = c()


for (rec in unique(data$Recording)) {
    
    if (rec %in% analysedRuns) {
        next
    }
        
    runs = getSameCondRuns(rec)
    
    for (runNum in runs) {
        analysedRuns[length(analysedRuns)+1] = runNum
    }
    
    run1 = data[data$Recording == runs[1], ]
    run2 = data[data$Recording == runs[2], ]
    run3 = data[data$Recording == runs[3], ]
    
    shortest = 120 * 15 #min(nrow(run1), nrow(run2), nrow(run3))
    
    run1 = head(run1, shortest)
    run2 = head(run2, shortest)
    run3 = head(run3, shortest)
    
    run1$setRun = 1
    run2$setRun = 2
    run3$setRun = 3

    run1$posDif = sqrt((run1$GreenX-run2$GreenX)^2 + (run1$GreenY-run2$GreenY)^2)
    run2$posDif = sqrt((run2$GreenX-run3$GreenX)^2 + (run2$GreenY-run3$GreenY)^2)
    run3$posDif = sqrt((run1$GreenX-run3$GreenX)^2 + (run1$GreenY-run3$GreenY)^2)
    
    totalDif = run1$posDif + run2$posDif + run3$posDif
    
    if (unique(run1$Type) != "angle") {
        next
    }
    
    
    difSum = c(0)
    
    for (elem in totalDif) {
        difSum[length(difSum)+1] = difSum[length(difSum)]+elem/360
    }
    
    difSum = head(difSum, length(difSum)-1)
    
    
    plot(run1$Time, difSum, cex=0.1, ylim=c(0, 12.5), type="l",
         ylab="Cumulative Chaos (m²·s)", xlab="Time (s)",
         col=rgb(unique(run1$InitAngle) / 180,
                 0.1,
                 1-unique(run1$InitAngle) / 180))
    
    par(new=T)
    
    difSet[length(difSet)+1] = sum(totalDif)
    angSet[length(angSet)+1] = unique(run1$InitAngle)
}
```

```{r, echo=FALSE}

energySet = sin(angSet * (3.14159 / 360))

difSet = difSet / 360

#plot(angSet, difSet, xlim=c(0,180), ylim=c(0, 220000), xlab="Angle (°)", ylab="Chaos over 15 secconds (m²·s)")

plot(energySet, difSet, xlim=c(0,1), ylim=c(0, 12.5), cex=0.8,
     xlab="Initial Energy (J/kg/m)", 
     ylab="Chaos over 15 seconds (m²·s)")


# Uncertainty in node position:
err = c()

for (i in 1:length(angSet)) {
    if (i < 4) {
        err[i] = sin(angSet[i] * (3.14159 / 90))
    } else {
        err[i] = 0.3
    }
}


segments(energySet, difSet-err, energySet, difSet+err)

epsilon <- 0.006
segments(energySet-epsilon, difSet-err, energySet+epsilon, difSet-err)
segments(energySet-epsilon, difSet+err, energySet+epsilon, difSet+err)



linearModelEnergyDif = lm(formula = difSet ~ 0+energySet)


abline(linearModelEnergyDif, col="blue")

```

```{r, echo=FALSE}
summary(linearModelEnergyDif)
```

## Conclusion:

Address the aim/question based on evidence from your results.\
Any limitations (with uncertainties) and future work\
A few sentences.

```{r}

```

<br>

# References
